---
title: "Cyclistic Data Project"
author: "Samuel Maduka"
date: "12/16/2021"
output:
  pdf_document: default
  html_document: default
---

##Load the packages
```{r}
library("tidyverse")
library("lubridate")
```


##Set working directory 
```{r}
getwd()
setwd("/Users/Probook 4730s/Desktop/tripdata")
```


## Upload datasets (csv files) here
```{r}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read.csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read.csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read.csv("Divvy_Trips_2020_Q1.csv")
```


#WRANGLE DATA AND COMBINE INTO A SINGLE FILE

##Compare column names of each files, column names need to match perfectly 
```{r}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)

```

##Rename columns to be consistent with q1_2020
```{r}
q4_2019 <- rename(q4_2019
                  ,ride_id = trip_id
                  ,rideable_type = bikeid 
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name 
                  ,end_station_id = to_station_id 
                  ,member_casual = usertype)
q3_2019 <- rename(q3_2019
                  ,ride_id = trip_id
                  ,rideable_type = bikeid 
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name 
                  ,end_station_id = to_station_id 
                  ,member_casual = usertype)
q2_2019 <- rename(q2_2019
                  ,ride_id = "01 - Rental Details Rental ID"
                  ,rideable_type = "01 - Rental Details Bike ID" 
                  ,started_at = "01 - Rental Details Local Start Time"  
                  ,ended_at = "01 - Rental Details Local End Time"  
                  ,start_station_name = "03 - Rental Start Station Name" 
                  ,start_station_id = "03 - Rental Start Station ID"
                  ,end_station_name = "02 - Rental End Station Name" 
                  ,end_station_id = "02 - Rental End Station ID"
                  ,member_casual = "User Type")
```



##Inspect the dataframes
```{r}
str(q1_2020)
str(q2_2019)
str(q3_2019)
str(q4_2019)
```


##Convert ride_id and rideable_type to character so they can stack
```{r}
q4_2019 <- mutate(q4_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
q3_2019 <- mutate(q3_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
q2_2019 <- mutate(q2_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type))
```


##Convert all started_at and ended_at columns from character to datetime using the POSIXct function
```{r}
q4_2019$started_at <- as.POSIXct(q4_2019$started_at)
q4_2019$ended_at <- as.POSIXct(q4_2019$ended_at)
q3_2019$started_at <- as.POSIXct(q3_2019$started_at)
q3_2019$ended_at <- as.POSIXct(q3_2019$ended_at)
q1_2020$started_at <- as.POSIXct(q1_2020$started_at)
q1_2020$ended_at <- as.POSIXct(q1_2020$ended_at)
```


##Stack individual quarter's data frames into one
```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```


##Remove data fields that was dropped in 2020 
```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```



##Inspecting the new table
```{r}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```


##Adjust the member_casual column to have just two labels instead of four
###To check the four labels
```{r}
table(all_trips$member_casual)
all_trips <- all_trips %>% 
  mutate(member_casual = recode(member_casual, "Subscriber" = "member", "Customer" = "casual"))
```


##Add columns that list the date, month, day, and year of each ride
```{r}
all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day <- format(as.Date(all_trips$date), "%A")
```


##We add ride_length calculation
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at)
```

##Inspect the data
```{r}
str(all_trips)
colnames(all_trips)
```


##We convert ride_length from factor to numeric to enable us do calculations
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```


## The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
## Create a new version of the all_trips dataframe since we removing data
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```


#Conduct Descriptive Analysis
```{r}
mean(all_trips_v2$ride_length) #Average ride
median(all_trips_v2$ride_length) #Midpoint number in ascending order of ride_length
max(all_trips_v2$ride_length) #Longest ride
min(all_trips_v2$ride_length) #Shortest Ride
summary(all_trips_v2$ride_length) #Gather all in a table
```


##Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```


## See the average ride time by each day for members vs casual users
```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day, FUN = mean)
```


## Notice that the days of the week are out of order. Let's fix that.
```{r}
all_trips_v2$day <- ordered(all_trips_v2$day, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


##Analyze ridership data by type and day
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% #Creates weekday fields using wday()
  group_by(member_casual, weekday) %>% #group by usertype and weekday
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% #calculate number of rides and average duration
  arrange(member_casual, weekday) %>% #Sort by usertype and weekday
  #Lets Visualize the number of rides by rider type 
  ggplot(aes(x=weekday, y=number_of_rides, fill = member_casual)) + geom_col(position = "dodge")
```


##Create viz for average duration  
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% #Creates weekday fields using wday()
  group_by(member_casual, weekday) %>% #group by usertype and weekday
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% #calculate number of rides and average duration
  arrange(member_casual, weekday) %>% #Sort by usertype and weekday
  #Lets Visualize the number of rides by rider type 
  ggplot(aes(x=weekday, y=average_duration, fill = member_casual)) + geom_col(position = "dodge")
```




